\section{Verifying the implementation} \label{q3}
In \crefrange{block}{MPIalg} verification has been added in the form of \textit{ensures} and \textit{requires}.
The verification guarantees that the work-items do not write to the same memory location.
This is required to ensure the determinism of the algorithm, so that the result is consistent.
As explained in \cref{backver}, permission-based separation logic is used to state which work-item has access to what resources.

Most of the lines are trivial or explained in \cref{algorithms}.
Non-trivial lines will be explained in this section.
The OpenCL algorithms are executed as a single group, so the work-group verification is equal to the kernel verification
Therefore, the work-group verification has been left out.

\subsection{OpenCL Algorithm \ref*{block}} \label{ver:block}
In \cref{block}, \crefrange{block:barrier:requires}{block:barrier} describe how the barrier distributes the read and write permissions among the work-items.
\Cref{block:barrier:requires} reclaims all permissions the work-items have on $d$.
The following line redistributes the permissions, and \cref{block:barrier:global} describes how the permissions are distributed over the kernel.
Since the difference of the $x$s between two consecutive threads is 2, no threads should have read permissions on a cell with write permissions of another work-item.

\Cref{block:firstcell,block:lastcell} refer to the fact that the first and last cells in the array of diagonals are not edited, as \Cref{singlevar} illustrates.

\subsection{OpenCL Algorithm \ref*{begin}} \label{ver:begin}
The explanation of the previous section also holds up for \cref{begin}, but a few line have been added.
\Cref{begin:dall,begin:did,begin:dwidth} are required in this algorithm to enable the writes in \crefrange{begin:fillbegin}{begin:fillend}.
\Cref{begin:barrier:did,begin:barrier:dwidth} reclaim those permissions.
\Crefrange{begin:barrier:requires}{begin:barrier:global} of \cref{begin} are equivalent to \crefrange{block:barrier:requires}{block:barrier:global} from \cref{block}.
The if statement on \cref{begin:ifbegin} helps with the enforcement of permissions, as less work-items process cells at the same time.

\subsection{OpenCL Algorithm \ref*{fill_column}} \label{ver:fill_column}
Verification of \cref{fill_column} is trivial, as no work-item requires read permissions and no write permissions overlap.
\Cref{fill_column:dall,fill_column:did} describe what write permissions are required.

\input{parts/graphs}

\subsection{MPI Algorithm \ref*{MPIalg}} \label{ver:MPIalg}
As the nodes in a cluster only communicate the right most columns of blocks, no concurrent variable manipulation can occur.
Therefore, permission-based separation logic is not required for this algorithm.
There are no circular dependencies, so no deadlocks can occur.
This cuts down the need for verification even further, leaving only trivial \textit{requires} and \textit{ensures} statements.

\subsection{Verification with VerCors} \label{ver:vercors}
Unfortunately, OpenCL is not sufficiently supported by VerCors to do automatic verification.
Rewriting the kernel in PVL, the native language of VerCors, does not help, as there is no support for kernel arguments.
In further research support could be built into VerCors, but in the mean time the manual verification provided in \crefrange{block}{MPIalg} will have to do.
